CF Registry Analysis - Lung Function

### 2025-12-05 Christine Chan/CF Registry/cf_registry_lung_function_analysis.qmd
fixed lung function analysis

```{r setup}
#| include: false
library(ggplot2)
library(gtsummary)
library(gt)
library(pROC)
library(ggrepel)
library(childsds)
library(labelled)
library(Hmisc)
library(lmerTest)
library(multcomp)
library(tidyverse)
home_dir <- switch(
  Sys.info()["sysname"],
  "Darwin" = "/Users/tim/Library/CloudStorage/OneDrive-TheUniversityofColoradoDenver/Vigers/CF/Christine Chan/CF Registry",
  "Windows" = "C:/Users/tim/OneDrive - The University of Colorado Denver/Vigers/CF/Christine Chan/CF Registry",
  "Linux" = "/home/tim/OneDrive/Vigers/CF/Christine Chan/CF Registry"
)
github_dir <- switch(
  Sys.info()["sysname"],
  "Darwin" = "/Users/timvigers/GitHub",
  "Windows" = "C:/Users/Tim/GitHub",
  "Linux" = "/home/tim/GitHub"
)
knitr::opts_knit$set(root.dir = home_dir)
set.seed(1017)
```

# Participant characteristics at CFRD diagnosis by diagnostic method

@@

data_dict <- readxl::read_excel("./Data_Raw/Copy of Codes for CFFPR_2018.xlsx")
data_dict$var <- paste(data_dict$field_name, data_dict$code_value, sep = "_")

@@

# Variable formatting
# If A_supplement_feeding is 0, then A_feeding2 is also 0 because they are not
# on any supplemental feeding
# Include all kinds of tube feeds
annual$A_feeding2[annual$A_supplement_feeding == 0] <- 0
annual$A_feeding3[annual$A_supplement_feeding == 0] <- 0
annual$A_feeding4[annual$A_supplement_feeding == 0] <- 0
tube_fields <- c("A_feeding2", "A_feeding3", "A_feeding4")
annual$tube_feeding <-
  apply(annual[, tube_fields], 1, function(r) {

@@

annual$bmi_adult = annual$bmi
annual$bmi_adult[annual$Age_YrEnd <= 20] = NA
annual$bmi_perc = annual$bmi_perc
annual$bmi_perc[annual$Age_YrEnd > 20] = NA

@@

# Age groups 10 - 18 inclusive, 19 - 30 inclusive, >= 31
annual$age_group <- cut(
  annual$Age_YrEnd,
  breaks = c(-Inf, 10, 19, 31, Inf),

@@

annual$cfrd_treatment_options4[annual$cfrd_treatment_options1 == 4] <- NA

@@

  "tube_feeding",
  "cfrd_treatment",
  "cfrd_treatment_options1",

@@

# FEV1 to categorical
annual$ppFEV1_cat <- cut(
  annual$A_FEV1_pct_predicted,
  breaks = c(0, 40, 80, 100, Inf),
annual$ppFEV1_cat <- factor(
  annual$ppFEV1_cat,
  levels = c("[100,Inf)", "[80,100)", "[40,80)", "[0,40)")

@@

# Add formatted columns for survival analysis
annual$tstop <- ymd(paste0(annual$ReviewYear, "-12-31"))
annual$tstart <- ymd(paste0(annual$ReviewYear, "-01-01"))
annual <- annual |>
  mutate(entry_date = tstart[1]) |>
  ungroup()
# Remove those under 10
annual$age_group <- relevel(annual$age_group, ref = "[10,19)")
annual <- annual |> filter(Age_YrEnd != 0, Age_YrEnd >= 10)
annual$age_group <- droplevels(annual$age_group)
# Get first CFRD again for people who were diagnosed before age 10
annual$first_cfrd = NULL
# Number of visits
annual <- annual |> group_by(eDWID) |> mutate(years_with_data = n())
# List of covariates
vars <- c(
  "age_group",
  "Gender",
  "bmi_perc",
  "bmi_cat",
  "MutClass",
  "Race",
  "Hispanicrace",
  "A_IsOnEnzymes",
  "ppFEV1_cat",
  "tube_feeding",
  "NumPulmExacerbation",
  "cfrd_status_annual",
  "liver_disease"
# Nice labels for plots and tables
labels <- list(
  cfrd_status_annual = "CFRD Status",
  MutClass = "Mutation class",
  Race = "Race",
  Hispanicrace = "Hispanic",
  Gender = "Sex",
  A_IsOnEnzymes = "Pancreatic enzyme use",
  ppFEV1_cat = "FEV1 percent predicted category",
  A_FEV1_pct_predicted = "FEV1 percent predicted",
  A_FVC_pct_predicted = "FVC percent predicted",
  tube_feeding = "Enteric feedings",
  liver_disease = "Liver disease",
  A_corticosteroids1 = "Use of corticosteroids",
  NumPulmExacerbation = "Admissions for PEx",
  age_group = "Age group",
  height_perc = "Height percentile",
  A_Vx770 = "Use of CFTR modulator Vx770",
  A_VX809comb = "Use of CFTR modulator Vx809 combination",
  A_pseudomonasaeruginosa = "Pseudomonas aeruginosa",
  A_fungalyeast1 = "Aspergillus (any species)",
  A_mycocultureresults3 = "Microorganisms in myco culture",
  A_cultureresults3 = "Microorganisms in culture",
  eDWID = "ID",
  tstart = "Interval start",
  tstop = "Interval stop",
  Age_YrEnd = "Age",
  cfrd_treatment = "Prescribed treatment for CFRD",
  cfrd_treatment_options1 = "Dietary change used to treat CFRD during the year",
  cfrd_treatment_options2 = "Oral hypoglycemic agents used to treat CFRD during the year",
  cfrd_treatment_options3 = "Intermittent insulin used to treat CFRD during the year",
  cfrd_treatment_options4 = "Chronic insulin used to treat CFRD during the year",
  years_with_data = "Years With Data"
label(annual[, which(colnames(annual) %in% names(labels))]) <-
  labels[colnames(annual)[which(colnames(annual) %in% names(labels))]]
# Per reviewer, out of X individuals with a diagnosis of CFRD, X number had an
# OGTT (and X had an A1c, but no OGTT) that year or the year prior.
# This is super slow but idc idc
cfrd_only <- apply(first_cfrd, 1, function(r) {

@@

  return(c(id, year, a1c, ogtt0hr, ogtt2hr))

@@

colnames(cfrd_only) <- c("id", "year", "a1c", "ogtt0hr", "ogtt2hr")
cfrd_only <- left_join(
  cfrd_only,
  annual |> mutate(eDWID = as.numeric(eDWID)),
  by = join_by(id == eDWID, year == ReviewYear)

@@

@@

cfrd_only |>
  select(
    a1c_ogtt,
    Age_YrEnd,
    A_weight,
    A_height,
    bmi_adult,
    bmi_cat,
    A_FEV1_pct_predicted,
    A_FVC_pct_predicted,
    Race,
    Hispanicrace,
    MutClass,
    a1c,
    A_ogttfast_bloodglucose,
    A_twohour_bloodglucose,
    A_IsOnEnzymes,
    liver_disease,
    A_pseudomonasaeruginosa,
    A_fungalyeast1,
    A_mycocultureresults3,
    A_cultureresults3,
    cfrd_treatment,
    cfrd_treatment_options1,
    cfrd_treatment_options2,
    cfrd_treatment_options3,
    cfrd_treatment_options4,
    years_with_data
  ) |>
  tbl_summary(
    by = a1c_ogtt,
    missing_text = "Missing",
    statistic = list(all_continuous() ~ "{mean} ({sd})")
  ) |>
  add_overall() |>
  add_p(
    test = list(
      all_continuous() ~ "oneway.test",
      "years_with_data" ~ "kruskal.test"

@@

  ) |>
  modify_caption("**All Groups**")
cfrd_only |>
  filter(a1c_ogtt %in% c("OGTT", "HbA1c Only")) |>
  mutate(a1c_ogtt = droplevels(a1c_ogtt)) |>
  select(
    a1c_ogtt,
    Age_YrEnd,
    A_weight,
    A_height,
    A_bmivalue,
    A_bmipercentile,
    A_FEV1_pct_predicted,
    A_FVC_pct_predicted,
    Race,
    Hispanicrace,
    MutClass,
    a1c,
    A_ogttfast_bloodglucose,
    A_twohour_bloodglucose,
    A_IsOnEnzymes,
    liver_disease,
    A_pseudomonasaeruginosa,
    A_fungalyeast1,
    A_mycocultureresults3,
    A_cultureresults3,
    cfrd_treatment,
    cfrd_treatment_options1,
    cfrd_treatment_options2,
    cfrd_treatment_options3,
    cfrd_treatment_options4,
    years_with_data
  ) |>
  tbl_summary(
    by = a1c_ogtt,
    missing_text = "Missing",
    statistic = list(all_continuous() ~ "{mean} ({sd})")
  ) |>
  add_overall() |>
  add_p(test = list(all_continuous() ~ "oneway.test")) |>
  modify_caption("**OGTT vs. HbA1c Only**")

@@

## FEV1 trends by diagnosis method

@@

```{r}
#| warning: false
#| message: false
# Plot by diagnosis type
mod_df = cfrd_only |>
  select(id, a1c_ogtt, Age_YrEnd) |>
  rename(eDWID = id)
mod_df = left_join(
  annual |> select(-Age_YrEnd) |> mutate(eDWID = as.numeric(eDWID)),
  mod_df,
  by = join_by(eDWID)
)
mod_df = mod_df |>
  drop_na(first_cfrd) |>
  group_by(eDWID) |>
  mutate(years_to_cfrd = ReviewYear - first_cfrd)
var_label(mod_df$years_to_cfrd) = "Years to CFRD"
var_label(mod_df$Age_YrEnd) = "Age at Diagnosis"
ggplot(
  mod_df |> filter(!is.na(a1c_ogtt)),
  aes(
    y = A_FEV1_pct_predicted,
    x = years_to_cfrd,
    color = a1c_ogtt,
    group = a1c_ogtt
  )
) +
  xlab("Years from CFRD") +
  xlim(c(-5, 5)) +
  geom_point(alpha = 0.1, shape = ".") +
  geom_smooth(formula = y ~ s(x, bs = "cr")) +
  theme_bw() +
  theme(legend.title = element_blank())
```

### Piecewise linear model results

```{r}
# Fit model - adjust for age at dx, sex, genotype (severe vs mild), pancreatic
# insufficiency, age at CF diagnosis (do we know this?), number of clinic
# visits/year, use of CFTR modulators
mod_df$x = mod_df$years_to_cfrd * (mod_df$years_to_cfrd >= 0)
label(mod_df$x) = "Change in Slope Post-CFRD"
label(mod_df$years_to_cfrd) = "Slope Pre-CFRD"
label(mod_df$a1c_ogtt) = "Diagnosis Method"
label(mod_df$Age_YrEnd) = "Age at Diagnosis"
# Write data so Christine can see missing data
t = mod_df |>
  select(
    eDWID,
    years_to_cfrd,
    A_FEV1_pct_predicted,
    A_FVC_pct_predicted,
    a1c_ogtt,
    Age_YrEnd,
    Gender,
    MutClass,
    A_IsOnEnzymes,
    A_Vx770,
    A_VX809comb,
    cfrd_treatment,
    cfrd_treatment_options1,
    cfrd_treatment_options2,
    cfrd_treatment_options3,
    cfrd_treatment_options4
  ) |>
  drop_na(A_FEV1_pct_predicted)
write.csv(t, "./Data_Clean/fev1_model_data.csv", row.names = F, na = "")
# Fit the model
m <- lmer(
  A_FEV1_pct_predicted ~ years_to_cfrd *
    a1c_ogtt +
    x * a1c_ogtt +
    Age_YrEnd +
    Gender +
    MutClass +
    A_IsOnEnzymes +
    A_Vx770 +
    A_VX809comb +
    (1 | eDWID),
  data = mod_df
)
# Population-averaged plot
pop_dat <- data.frame(expand.grid(
  years_to_cfrd = -5:5,
  a1c_ogtt = unique(mod_df$a1c_ogtt),
  Age_YrEnd = mean(mod_df$Age_YrEnd, na.rm = T),
  Gender = "M",
  MutClass = "1-3",
  A_IsOnEnzymes = "Yes",
  A_Vx770 = "No",
  A_VX809comb = "No"
))
pop_dat$x <- (pop_dat$years_to_cfrd >= 0) * pop_dat$years_to_cfrd
pred <- predict(m, re.form = NA, newdata = pop_dat, se.fit = T)
pop_dat$pred <- pred$fit
pop_dat$ll <- pop_dat$pred - (1.96 * pred$se.fit)
pop_dat$ul <- pop_dat$pred + (1.96 * pred$se.fit)
ggplot(
  pop_dat,
  aes(x = years_to_cfrd, y = pred, group = a1c_ogtt, colour = a1c_ogtt)
) +
  geom_point() +
  geom_errorbar(aes(ymin = ll, ymax = ul), width = 0.1) +
  geom_line() +
  xlab("Years from CFRD") +
  ylab("FEV1 % Predicted") +
  theme_classic(base_size = 15)
# Contrasts
neither_pre <- rep(0, nrow(summary(m)$coefficients))
neither_change <- rep(0, nrow(summary(m)$coefficients))
neither_pre[which(rownames(summary(m)$coefficients) == "years_to_cfrd")] = 1
neither_change[which(rownames(summary(m)$coefficients) == "x")] = 1
a1c_pre <- rep(0, nrow(summary(m)$coefficients))
a1c_change <- rep(0, nrow(summary(m)$coefficients))
a1c_pre[which(
  rownames(summary(m)$coefficients) %in%
    c("years_to_cfrd", "years_to_cfrd:a1c_ogttHbA1c Only")
)] = 1
a1c_change[which(
  rownames(summary(m)$coefficients) %in%
    c("x", "a1c_ogttHbA1c Only:x")
)] = 1
ogtt_pre <- rep(0, nrow(summary(m)$coefficients))
ogtt_change <- rep(0, nrow(summary(m)$coefficients))
ogtt_pre[which(
  rownames(summary(m)$coefficients) %in%
    c("years_to_cfrd", "years_to_cfrd:a1c_ogttOGTT")
)] = 1
ogtt_change[which(
  rownames(summary(m)$coefficients) %in%
    c("x", "a1c_ogttOGTT:x")
)] = 1
contr <- rbind(
  "HbA1c Only Pre-CFRD" = a1c_pre,
  "HbA1c Only Post-CFRD" = a1c_pre + a1c_change,
  "OGTT Pre-CFRD" = ogtt_pre,
  "OGTT Post-CFRD" = ogtt_pre + ogtt_change,
  "Neither Pre-CFRD" = neither_pre,
  "Neither Post-CFRD" = neither_pre + neither_change
)
change_mod_slopes <- glht(m, linfct = contr)
change_mod_slopes_ci <- confint(change_mod_slopes, calpha = univariate_calpha())
change_mod_slopes_ci$confint |>
  as.data.frame() |>
  gt(caption = "Slopes by Diagnosis Method", rownames_to_stub = T) |>
  fmt_number(decimals = 2)
# Model table
m |> tbl_regression() |> add_n(location = "level")
```

### Categorical model results

## FVC trends by diagnosis method

@@

```{r} 
ggplot(
  mod_df |> filter(!is.na(a1c_ogtt)),
  aes(
    y = A_FVC_pct_predicted,
    x = years_to_cfrd,
    color = a1c_ogtt,
    group = a1c_ogtt
  )
) +
  xlab("Years from CFRD") +
  xlim(c(-5, 5)) +
  geom_point(alpha = 0.1, shape = ".") +
  geom_smooth(formula = y ~ s(x, bs = "cr")) +
  theme_bw() +
  theme(legend.title = element_blank())
```

### Piecewise linear model results

```{r}
# Fit model - adjust for age at dx, sex, genotype (severe vs mild), pancreatic
# insufficiency, age at CF diagnosis (do we know this?), number of clinic
# visits/year, use of CFTR modulators
mod_df$x = mod_df$years_to_cfrd * (mod_df$years_to_cfrd >= 0)
label(mod_df$x) = "Change in Slope Post-CFRD"
label(mod_df$years_to_cfrd) = "Slope Pre-CFRD"
label(mod_df$a1c_ogtt) = "Diagnosis Method"
label(mod_df$Age_YrEnd) = "Age at Diagnosis"
m <- lmer(
  A_FVC_pct_predicted ~ years_to_cfrd *
    a1c_ogtt +
    x * a1c_ogtt +
    Age_YrEnd +
    Gender +
    MutClass +
    A_IsOnEnzymes +
    A_Vx770 +
    A_VX809comb +
    (1 | eDWID),
  data = mod_df
)
# Population-averaged plot
pop_dat <- data.frame(expand.grid(
  years_to_cfrd = -5:5,
  a1c_ogtt = unique(mod_df$a1c_ogtt),
  Age_YrEnd = mean(mod_df$Age_YrEnd, na.rm = T),
  Gender = "M",
  MutClass = "1-3",
  A_IsOnEnzymes = "Yes",
  A_Vx770 = "No",
  A_VX809comb = "No"
))
pop_dat$x <- (pop_dat$years_to_cfrd >= 0) * pop_dat$years_to_cfrd
pred <- predict(m, re.form = NA, newdata = pop_dat, se.fit = T)
pop_dat$pred <- pred$fit
pop_dat$ll <- pop_dat$pred - (1.96 * pred$se.fit)
pop_dat$ul <- pop_dat$pred + (1.96 * pred$se.fit)
ggplot(
  pop_dat,
  aes(x = years_to_cfrd, y = pred, group = a1c_ogtt, colour = a1c_ogtt)
) +
  geom_point() +
  geom_errorbar(aes(ymin = ll, ymax = ul), width = 0.1) +
  geom_line() +
  xlab("Years from CFRD") +
  ylab("FVC % Predicted") +
  theme_classic(base_size = 15)
# Contrasts
neither_pre <- rep(0, nrow(summary(m)$coefficients))
neither_change <- rep(0, nrow(summary(m)$coefficients))
neither_pre[which(rownames(summary(m)$coefficients) == "years_to_cfrd")] = 1
neither_change[which(rownames(summary(m)$coefficients) == "x")] = 1
a1c_pre <- rep(0, nrow(summary(m)$coefficients))
a1c_change <- rep(0, nrow(summary(m)$coefficients))
a1c_pre[which(
  rownames(summary(m)$coefficients) %in%
    c("years_to_cfrd", "years_to_cfrd:a1c_ogttHbA1c Only")
)] = 1
a1c_change[which(
  rownames(summary(m)$coefficients) %in%
    c("x", "a1c_ogttHbA1c Only:x")
)] = 1
ogtt_pre <- rep(0, nrow(summary(m)$coefficients))
ogtt_change <- rep(0, nrow(summary(m)$coefficients))
ogtt_pre[which(
  rownames(summary(m)$coefficients) %in%
    c("years_to_cfrd", "years_to_cfrd:a1c_ogttOGTT")
)] = 1
ogtt_change[which(
  rownames(summary(m)$coefficients) %in%
    c("x", "a1c_ogttOGTT:x")
)] = 1
contr <- rbind(
  "HbA1c Only Pre-CFRD" = a1c_pre,
  "HbA1c Only Post-CFRD" = a1c_pre + a1c_change,
  "OGTT Pre-CFRD" = ogtt_pre,
  "OGTT Post-CFRD" = ogtt_pre + ogtt_change,
  "Neither Pre-CFRD" = neither_pre,
  "Neither Post-CFRD" = neither_pre + neither_change
)
change_mod_slopes <- glht(m, linfct = contr)
change_mod_slopes_ci <- confint(change_mod_slopes, calpha = univariate_calpha())
change_mod_slopes_ci$confint |>
  as.data.frame() |>
  gt(caption = "Slopes by Diagnosis Method", rownames_to_stub = T) |>
  fmt_number(decimals = 2)
# Model table
m |> tbl_regression() |> add_n(location = "level")
```

>>>>

```{r setup}
#| include: false
library(ggplot2)
library(gtsummary)
library(gt)
library(pROC)
library(ggrepel)
library(childsds)
library(labelled)
library(Hmisc)
library(lmerTest)
library(emmeans)
library(multcomp)
library(tidyverse)
library(parallel)
library(knitr)
base_dir <- switch(
  Sys.info()["nodename"],
  "togo" = "/home/tvigers/Documents/Data",
  "Tims-MacBook-Air.local" = "/Users/tim/Library/CloudStorage/OneDrive-TheUniversityofColoradoDenver/Vigers",
  "Tims-Mac-mini.local" = "/Users/tim/Library/CloudStorage/OneDrive-TheUniversityofColoradoDenver/Vigers",
  "Mac" = "/Users/tim/Library/CloudStorage/OneDrive-TheUniversityofColoradoDenver/Vigers"
)
data_dir <- switch(
  Sys.info()["nodename"],
  "togo" = "/CF/Christine Chan/CF Registry",
  "Tims-MacBook-Air.local" = "/CF/Christine Chan/CF Registry",
  "Tims-Mac-mini.local" = "/CF/Christine Chan/CF Registry",
  "Mac" = "/CF/Christine Chan/CF Registry"
)
home_dir <- paste0(base_dir, data_dir)
knitr::opts_knit$set(root.dir = home_dir)
```

@@

@@

@@

annual$bmi_adult <- annual$bmi
annual$bmi_adult[annual$Age_YrEnd <= 20] <- NA
annual$bmi_perc <- annual$bmi_perc
annual$bmi_perc[annual$Age_YrEnd > 20] <- NA

@@

@@

annual$cfrd_treatment_options4[annual$cfrd_treatment_options4 == 4] <- NA

@@

@@

@@

# Check whether they had an OGTT and/or HbA1c the year of first CFRD or the year
# prior. This is super slow but idc idc.
cl <- makeCluster(type = "FORK", 8)
cfrd_only <- parApply(
  cl,
  first_cfrd,
  1,
  function(r) {

@@

    d <- annual[annual$eDWID == id & annual$ReviewYear %in% c(year, year + 1), ]
    treatment = any(
      d[, c(
      )] ==
        "Yes",
      na.rm = TRUE
    return(c(id, year, a1c, ogtt0hr, ogtt2hr, treatment))
stopCluster(cl)

@@

colnames(cfrd_only) <- c(
  "eDWID",
  "dx_year",
  "a1c",
  "ogtt0hr",
  "ogtt2hr",
  "treatment"

@@

# cfrd_only_store <- cfrd_only

@@

var_label(cfrd_only$a1c_ogtt) <- "Diagnosis Method"
cfrd_only$treatment = factor(
  cfrd_only$treatment,
  labels = c("No Treatment", "Treatment")

@@

var_label(cfrd_only$treatment) = "CFRD Treatment Started Within 1 Year?"
# Calculate number of visits (just for table 1)
n_vis = annual |>
  summarise(years_with_data = n())
cfrd_only <- left_join(cfrd_only, n_vis, by = join_by(eDWID))

@@

# Diagnosed with CFRD between 2004 - 2018

## Participant characteristics at diagnosis

```{r}
# Create a model dataframe
mod_df = inner_join(
  annual,
  cfrd_only |> filter(dx_year >= 2004),
  by = join_by(eDWID)
) |>
  drop_na(a1c_ogtt) |>
  mutate(years_to_cfrd = ReviewYear - first_cfrd) |>
  filter(years_to_cfrd <= 5 & years_to_cfrd >= -5) |>
  mutate(years_to_cfrd = factor(years_to_cfrd))
var_label(mod_df$years_to_cfrd) = "Years From CFRD Dx"
# Write data for Christine to look at
mod_df |>
  select(
    eDWID,
    Race,
    Hispanicrace,
    MutClass,
    ReviewYear,
    Age_YrEnd,
    CFRD,
    first_cfrd,
    age_dx,
    A_weight,
    A_height,
    bmi_adult,
    bmi_perc,
    bmi_cat,
    A_FEV1_pct_predicted,
    A_FVC_pct_predicted,
    a1c_ogtt,
    A_hgba1c,
    A_ogttfast_bloodglucose,
    A_twohour_bloodglucose,
    A_IsOnEnzymes,
    liver_disease,
    A_pseudomonasaeruginosa,
    A_fungalyeast1,
    A_mycocultureresults3,
    A_cultureresults3,
    treatment,
    cfrd_treatment_options2,
    cfrd_treatment_options3,
    cfrd_treatment_options4,
    years_with_data
  ) %>%
  write.csv(
    .,
    file = "./Data_Clean/fev1_model_data_2004.csv",
    row.names = FALSE,
    na = ""
  )
# Tables
mod_df |>
  filter(ReviewYear == first_cfrd) |>
  select(
    a1c_ogtt,
    age_dx,
    A_weight,
    A_height,
    bmi_adult,
    bmi_perc,
    bmi_cat,
    A_FEV1_pct_predicted,
    A_FVC_pct_predicted,
    Race,
    Hispanicrace,
    MutClass,
    a1c,
    A_ogttfast_bloodglucose,
    A_twohour_bloodglucose,
    A_IsOnEnzymes,
    liver_disease,
    A_pseudomonasaeruginosa,
    A_fungalyeast1,
    A_mycocultureresults3,
    A_cultureresults3,
    treatment,
    cfrd_treatment_options2,
    cfrd_treatment_options3,
    cfrd_treatment_options4,
    years_with_data
  ) |>
  tbl_summary(
    by = a1c_ogtt,
    missing_text = "Missing",
    statistic = list(all_continuous() ~ "{mean} ({sd})")
  ) |>
  add_overall() |>
  add_p(
    test = list(
      all_continuous() ~ "oneway.test",
      "years_with_data" ~ "kruskal.test"
    )
  ) |>
  modify_caption("**All Groups**")
mod_df |>
  filter(ReviewYear == first_cfrd, a1c_ogtt %in% c("OGTT", "HbA1c Only")) |>
  mutate(a1c_ogtt = droplevels(a1c_ogtt)) |>
  select(
    a1c_ogtt,
    age_dx,
    A_weight,
    A_height,
    A_bmivalue,
    A_bmipercentile,
    A_FEV1_pct_predicted,
    A_FVC_pct_predicted,
    Race,
    Hispanicrace,
    MutClass,
    a1c,
    A_ogttfast_bloodglucose,
    A_twohour_bloodglucose,
    A_IsOnEnzymes,
    liver_disease,
    A_pseudomonasaeruginosa,
    A_fungalyeast1,
    A_mycocultureresults3,
    A_cultureresults3,
    treatment,
    cfrd_treatment_options2,
    cfrd_treatment_options3,
    cfrd_treatment_options4,
    years_with_data
  ) |>
  tbl_summary(
    by = a1c_ogtt,
    missing_text = "Missing",
    statistic = list(all_continuous() ~ "{mean} ({sd})")
  ) |>
  add_overall() |>
  add_p(test = list(all_continuous() ~ "oneway.test")) |>
  modify_caption("**OGTT vs. HbA1c Only**")
```

## FEV1

@@

```{r}
#| warning: false
#| message: false
# Plot by diagnosis type
ggplot(
  mod_df,
  aes(
    y = A_FEV1_pct_predicted,
    x = years_to_cfrd,
    color = a1c_ogtt,
    group = a1c_ogtt
  )
) +
  xlab("Years from CFRD") +
  geom_smooth() +
  theme_bw()
```

### Linear model results

#### Not adjusted for treatment status

Adjusted for age at diagnosis, sex, mutation class, and pancreatic enzyme use.

```{r}
# Fit the model
m <- lmer(
  A_FEV1_pct_predicted ~ years_to_cfrd *
    a1c_ogtt +
    age_dx +
    Gender +
    MutClass +
    A_IsOnEnzymes +
    (1 | eDWID),
  data = mod_df
)
t3 = car::Anova(m, type = 'III', contrasts = "contr.sum")
emm = emmeans(m, specs = ~ years_to_cfrd + a1c_ogtt, pbkrtest.limit = 1e5)
emm = data.frame(emm)
ggplot(
  emm,
  aes(x = years_to_cfrd, y = emmean, group = a1c_ogtt, color = a1c_ogtt)
) +
  geom_point() +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.25) +
  geom_line() +
  xlab("Years From CFRD Dx") +
  ylab("Estimated Mean ppFEV1") +
  labs(color = "Diagnosis Method") +
  theme_bw()
# Model tables
kable(emm, digits = 3)
m |> tbl_regression() |> add_n(location = "level")
t3
```

The type 3 ANOVA indicates a significant interaction between time and diagnosis method (p =`r format.pval(t3[nrow(t3),"Pr(>Chisq)"], eps = 0.001)`).

#### Adjusted for treatment status

Adjusted for age at diagnosis, sex, mutation class, pancreatic enzyme use, and CFRD treatment.

```{r}
# Fit the model
m <- lmer(
  A_FEV1_pct_predicted ~ years_to_cfrd *
    a1c_ogtt +
    age_dx +
    Gender +
    MutClass +
    A_IsOnEnzymes +
    treatment +
    (1 | eDWID),
  data = mod_df
)
t3 = car::Anova(m, type = 'III', contrasts = "contr.sum")
emm = emmeans(m, specs = ~ years_to_cfrd + a1c_ogtt, pbkrtest.limit = 1e5)
emm = data.frame(emm)
ggplot(
  emm,
  aes(x = years_to_cfrd, y = emmean, group = a1c_ogtt, color = a1c_ogtt)
) +
  geom_point() +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.25) +
  geom_line() +
  xlab("Years From CFRD Dx") +
  ylab("Estimated Mean ppFEV1") +
  labs(color = "Diagnosis Method") +
  theme_bw()
# Model tables
kable(emm, digits = 3)
m |> tbl_regression() |> add_n(location = "level")
t3
```

The type 3 ANOVA indicates a significant interaction between time and diagnosis method (p =`r format.pval(t3[nrow(t3),"Pr(>Chisq)"], eps = 0.001)`).

## FVC

@@

```{r}
#| warning: false
#| message: false
# Plot by diagnosis type
ggplot(
  mod_df,
  aes(
    y = A_FVC_pct_predicted,
    x = years_to_cfrd,
    color = a1c_ogtt,
    group = a1c_ogtt
  )
) +
  xlab("Years from CFRD") +
  geom_smooth() +
  theme_bw()
```

### Linear model results

#### Not adjusted for treatment status

Adjusted for age at diagnosis, sex, mutation class, and pancreatic enzyme use.

```{r}
# Fit the model
m <- lmer(
  A_FVC_pct_predicted ~ years_to_cfrd *
    a1c_ogtt +
    age_dx +
    Gender +
    MutClass +
    A_IsOnEnzymes +
    (1 | eDWID),
  data = mod_df
)
t3 = car::Anova(m, type = 'III', contrasts = "contr.sum")
emm = emmeans(m, specs = ~ years_to_cfrd + a1c_ogtt, pbkrtest.limit = 1e5)
emm = data.frame(emm)
ggplot(
  emm,
  aes(x = years_to_cfrd, y = emmean, group = a1c_ogtt, color = a1c_ogtt)
) +
  geom_point() +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.25) +
  geom_line() +
  xlab("Years From CFRD Dx") +
  ylab("Estimated Mean ppFVC") +
  labs(color = "Diagnosis Method") +
  theme_bw()
# Model tables
kable(emm, digits = 3)
m |> tbl_regression() |> add_n(location = "level")
t3
```

The type 3 ANOVA indicates a significant interaction between time and diagnosis method (p =`r format.pval(t3[nrow(t3),"Pr(>Chisq)"], eps = 0.001)`).

#### Adjusted for treatment status

Adjusted for age at diagnosis, sex, mutation class, pancreatic enzyme use, and CFRD treatment.

```{r}
# Fit the model
m <- lmer(
  A_FVC_pct_predicted ~ years_to_cfrd *
    a1c_ogtt +
    age_dx +
    Gender +
    MutClass +
    A_IsOnEnzymes +
    treatment +
    (1 | eDWID),
  data = mod_df
)
t3 = car::Anova(m, type = 'III', contrasts = "contr.sum")
emm = emmeans(m, specs = ~ years_to_cfrd + a1c_ogtt, pbkrtest.limit = 1e5)
emm = data.frame(emm)
ggplot(
  emm,
  aes(x = years_to_cfrd, y = emmean, group = a1c_ogtt, color = a1c_ogtt)
) +
  geom_point() +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.25) +
  geom_line() +
  xlab("Years From CFRD Dx") +
  ylab("Estimated Mean ppFVC") +
  labs(color = "Diagnosis Method") +
  theme_bw()
# Model tables
kable(emm, digits = 3)
m |> tbl_regression() |> add_n(location = "level")
t3
```

The type 3 ANOVA indicates a significant interaction between time and diagnosis method (p =`r format.pval(t3[nrow(t3),"Pr(>Chisq)"], eps = 0.001)`).

# Diagnosed with CFRD between 2010 - 2018

```{r}
# Create a model dataframe
mod_df = inner_join(
  annual,
  cfrd_only |> filter(dx_year >= 2010),
  by = join_by(eDWID)
) |>
  drop_na(a1c_ogtt) |>
  mutate(years_to_cfrd = ReviewYear - first_cfrd) |>
  filter(years_to_cfrd <= 5 & years_to_cfrd >= -5) |>
  mutate(years_to_cfrd = factor(years_to_cfrd))
var_label(mod_df$years_to_cfrd) = "Years From CFRD Dx"
```

## Participant characteristics at diagnosis
